---
title: "BDA project report"
author: "Amanda Aarnio, Anni Niskanen, Antti Huttunen"
date: "`r Sys.Date()`"
output: pdf_document
---

# Introduction 
```{r, warning=FALSE, message=FALSE}
set.seed(123)
library(cmdstanr)
library(Stat2Data)
data("FirstYearGPA")
```
\newpage
# Data and Problem 

```{r, warning=FALSE, message=FALSE}
male_white <- FirstYearGPA[FirstYearGPA$Male==1 & FirstYearGPA$White==1,]
male_non_white <- FirstYearGPA[FirstYearGPA$Male==1 & FirstYearGPA$White==0,]
female_white <- FirstYearGPA[FirstYearGPA$Male==0 & FirstYearGPA$White==1,]
female_non_white <- FirstYearGPA[FirstYearGPA$Male==0 & FirstYearGPA$White==0,]
```

```{r, warning=FALSE, message=FALSE}
data_hierarchical <- list(N1 = nrow(male_white),
                            N2 = nrow(male_non_white),
                            N3 = nrow(female_white),
                            N4 = nrow(female_non_white),
                            x1 = subset(male_white, select = c('HSGPA', 'SATV', 'SATM','HU','SS')),
                            x2 = subset(male_non_white, select = c('HSGPA', 'SATV', 'SATM','HU','SS')),
                            x3 = subset(female_white, select = c('HSGPA', 'SATV', 'SATM','HU','SS')),
                            x4 = subset(female_non_white, select = c('HSGPA', 'SATV', 'SATM','HU','SS')),
                            y1 = male_white$GPA,
                            y2 = male_non_white$GPA,
                            y3 = female_white$GPA,
                            y4 = female_non_white$GPA,
                            musigma = 100,
                            sigmasigma = 10)
data_pooled <- list(N = nrow(FirstYearGPA),
                    x = subset(FirstYearGPA, select = c('HSGPA', 'SATV', 'SATM','HU','SS')),
                    y = FirstYearGPA$GPA)
```

\newpage
# Models 

## Pooled model 

**Mathematical notation**

$$
\begin{aligned}
GPA_i &\sim N(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_1 \cdot HSGPA_i + \beta_2 \cdot SATV_i + \beta_3 \cdot SATM_i + \beta_4 \cdot HU_i + \beta_5 \cdot SS_i \\
\sigma &\sim N(0, 10) \\
\alpha &\sim N(0, 100) \\
\beta_k &\sim N(0, 100) \\
\end{aligned}
$$
```{r}
writeLines(readLines("pooled.stan"))
mod_pooled <- cmdstan_model("pooled.stan")
fit_pooled <- mod_pooled$sample(data_pooled, refresh = 2000, seed = 123)
```


### Hierarchical model 

**Mathematical notation**

$$
\begin{aligned}
GPA_{ij} &\sim N(\mu_{ij}, \sigma) \\
\mu_{ij} &= \alpha_j + \beta_{1j} \cdot HSGPA_i + \beta_{2j} \cdot SATV_i + \beta_{3j} \cdot SATM_i + \beta_{4j} \cdot HU_i + \beta_{5j} \cdot SS_i \\
\sigma &\sim N(0, 10) \\
\alpha_j &\sim N(\mu_{\alpha}, \sigma_{\alpha}) \\
\beta_{1j} &\sim N(\mu_{\beta_1}, \sigma_{\beta_1}) \\
\beta_{2j} &\sim N(\mu_{\beta_2}, \sigma_{\beta_2}) \\
\beta_{3j} &\sim N(\mu_{\beta_3}, \sigma_{\beta_3}) \\
\beta_{4j} &\sim N(\mu_{\beta_4}, \sigma_{\beta_4}) \\
\beta_{5j} &\sim N(\mu_{\beta_5}, \sigma_{\beta_5}) \\
\mu_{\alpha} &\sim N(0,100) \\
\sigma_{\alpha} &\sim N(0,10) \\
\mu_{\beta_k} &\sim N(0,100) \\
\sigma_{\beta_k} &\sim N(0,10) \\
\end{aligned}
$$

```{r}
writeLines(readLines("hierarchical.stan"))
mod_hierarchical <- cmdstan_model("hierarchical.stan")
```
```{r}
fit_hierarchical <-mod_hierarchical$sample(data_hierarchical, refresh = 2000, seed = 123)
```

\newpage
# Analysis and Results

## Converge diagnostics

### pooled model

```{r}
sum(fit_pooled$summary()$rhat > 1.01)
sum(fit_pooled$summary()$rhat < 0.99)
```
No divergence

### hierarchical model

```{r}
sum(fit_hierarchical$summary()$rhat > 1.01)
sum(fit_hierarchical$summary()$rhat < 0.99)
```
Divergence: 172 of 4000 (4%)

ESS:
To-do 
Check whar's good ess!?

Questions for TA:
- What does it mean that values are higher that 1.01?
- What's the difference between ess_bulk and ess_tail?
- What's ESS? We had difficulties to understand course book.

## Posterior predictive checks

## Prior sensitivity analysis

\newpage
# Discussion

\newpage
# Conclusion

\newpage
# Self-reflection




